Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-11 supplementary/ptm_sm.tex
Added sample size and updated model section

\subsection{Data structure of PTM quantification experiments.}
A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}$. To account for the underlying protein abundance, features corresponding to the unmodified peptides from the same protein are considered together, except those unmodified peptides containing a modified site to avoid the confounding effect due to the PTM. The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}^{\ast}$. \sfigref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein. Unmodified peptides from the same protein provide additional evidence on the underlying protein abundance, which needs to be integrated for PTM characterization. To address the goals of PTM characterization, statistical analysis needs to summarize values in this table using appropriate statistical models, translate the goal into a model-based quantity of interest, and draw inference (i.e., characterize the uncertainty) about the quantity.

\begin{figure}[h!]
\centering
\begin{footnotesize}
\[
\begin{array}{l l c c c c c c c c c} 
\toprule
 &  & \multicolumn{4}{c}{\text{Condition }1} & \dots & \multicolumn{4}{c}{\text{Condition }I} \\ 
\cmidrule{3-6} 
\cmidrule{8-11} 
 &  &  \text{Run }1 & \text{Run }2 & \dots & \text{Run }J & \dots & \text{Run }1 & \text{Run }2 & \dots & \text{Run }J \\
\midrule
\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
 & \text{Feature }2 & y_{112} & - & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }K & y_{11K} & y_{12K} & \dots & y_{1JK} & \dots & y_{I1K} & y_{I2K} & \dots & y_{IJK} \\
\midrule
\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
 & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} & \dots & y_{1J2}^{\ast} & \dots & - & - & \dots & -  \\
% & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }L & y_{11L}^{\ast} & y_{12L}^{\ast} & \dots & y_{1JL}^{\ast} & \dots & y_{I1L}^{\ast} & y_{I2L}^{\ast} & \dots & y_{IJL}^{\ast} \\
\bottomrule
\end{array}
\]
%\[
%\begin{array}{l l c c c c c} 
%\toprule
% &  & \multicolumn{2}{c}{\text{Condition }1} &  & \multicolumn{2}{c}{\text{Condition }2} \\ 
%\cmidrule{3-4} 
%\cmidrule{6-7} 
% &  &  \text{Run }1 & \text{Run }2 &  & \text{Run }1 & \text{Run }2 \\
%\midrule
%\multirow{4}{*}{\begin{turn}{90} {\text{Modified}} \end{turn}} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
%%\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
% & \text{Feature }2 & y_{112} & - &  & y_{212} & y_{222} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }5 & y_{115} & y_{125} &  & y_{215} & y_{225} \\
%\midrule
%\multirow{7}{*}{\begin{turn}{90} {\text{Unmodified}} \end{turn}} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} &  & y_{211}^{\ast} & y_{221}^{\ast} \\
% & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} &  & - & y_{222}^{\ast} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }7 & y_{117}^{\ast} & y_{127}^{\ast} &  & y_{217}^{\ast} & y_{227}^{\ast} \\
% & \text{Feature }8 & y_{118}^{\ast} & y_{128}^{\ast} &  & y_{218}^{\ast} & y_{228}^{\ast} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }11 & y_{1,1,11}^{\ast} & y_{1,2,11}^{\ast} &  & y_{2,1,11}^{\ast} & y_{2,2,11}^{\ast} \\
%\bottomrule
%\end{array}
%\]
\end{footnotesize}
\caption{Representation of the data of modified peptides at one site and unmodified peptides of the same protein, with $I$ conditions and $J$ replicate runs. Abundances of the PTM and protein are quantified by multiple spectral features (peptide ions, $K$ for modified peptides and $L$ for unmodified peptides). Some spectral features can be missing (shown as $-$), either randomly in individual runs or completely in certain conditions. In real practice, the number of runs can vary across conditions. \label{fig:dtable}}
\end{figure}

@@

Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i'$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. The $t$-test is typically performed based on the log of summarized value. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right).
\]
For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
%which is equivalent to take the difference of their log abundances. 
The quantity is denoted by $d_{ij}$ and is given by
\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}^{\ast}} \right).
\]
Alternatively, run-level summary to be described in \secref{sec:sum} can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
\[
\hat{\Delta} = \frac{1}{J} d_{i+} - \frac{1}{J} d_{i'+},
\]
where $d_{i+} = \sum_{j=1}^{J} d_{ij}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

@@

\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right).
\]

@@

\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}^{\ast}} \right).
\]

@@

\[
\hat{\Delta} = \frac{1}{J} d_{i+} - \frac{1}{J} d_{i'+},
\]

@@

To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors: modification, condition, run, and feature. As different degrees of variability are present in the feature intensities of modified and unmodified peptides, they are expressed by separate models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Statistical modeling and inference}
\label{sec:model}

The observed log-intensity of a modified peptide feature is denoted by $y_{ijk}$ and represented as
\[
y_{ijk} = \psi + C_{i} + R_{j(i)} + F_{k} + (R \times F)_{ijk},
\]
where the effects of condition and feature are modeled as fixed effects: 
\[
\sum_{i=1}^{I} C_{i} = 0, \quad \sum_{k=1}^{K} F_{k} = 0,
\]
and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
\[
R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2}), \qquad
(R \times F)_{ijk} = \epsilon_{ijk} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon}^{2}).
\]
Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{ijl}^{\ast}$ and represented as 
\[
y_{ijl}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast} + F_{l}^{\ast} + (R \times F)_{ijl}^{\ast}, 
\]
where the effects of condition and feature are modeled as fixed effects: 
\[
\sum_{i=1}^{I} C_{i}^{\ast} = 0, \quad \sum_{l=1}^{L} F_{l}^{\ast} = 0,
\]
and 
\[
R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}), \qquad
(R \times F)_{ijl}^{\ast} = \epsilon_{ijl}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
\]

@@

Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a}, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish. \cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Model-based inference of the underlying abundance}
\label{sec:infer}

The PTM abundance in each run is represented as 
\[
\hat{y}_{ij} = \psi + C_{i} + R_{j(i)},
\]
where $\sum_{i=1}^{I} C_{i} = 0$, $R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$. 
Similarly, the protein abundance in each run is expressed as
\[
\hat{y}_{ij} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
\]
where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$. The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}$ and $\mu_{i}^{\ast}$, respectively, and the values are estimated as: 
\begin{align*}
\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+} \\
\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast},
\end{align*}
where the standard errors of the estimates are $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}^{\ast}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
%\[
%\left( \frac{\hat{\sigma}_{\gamma}^{2}}{J} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
%\]
%\[
%\left( \frac{1}{J} \hat{\sigma}_{\gamma}^{2} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
%\]
Based on the estimates $\hat{\mu}_{i}$ and $\hat{\mu}_{i}^{\ast}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i} - \hat{\mu}_{i}^{\ast})$ and the standard error of the estimate is 
\[
\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
\]

@@

With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
%Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
\begin{align*}
H_{0}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &\neq 0
\end{align*}
The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right],
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma}^{4}}{\mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{ \mathrm{df}(\gamma^{\ast})} \right).
\]
A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.

@@

The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma}^{2}$ and $\sigma_{\gamma^{\ast}}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively. The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations. With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}. Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
\[
\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]
\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
\]
where 
\[
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
\]
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs. More details can be found in~\cite{oberg_vitek_09a}.

@@

The log-fold change for adjusted PTM abundance is estimated as the average over mixtures and technical replicates

>>>>

\subsection{Data structure of PTM quantification experiments.}
A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. To account for the underlying protein abundance, features corresponding to the unmodified peptides from the same protein are considered together, except those unmodified peptides containing a modified site to avoid the confounding effect due to the PTM. The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. \sfigref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein. Unmodified peptides from the same protein provide additional evidence on the underlying protein abundance, which needs to be integrated for PTM characterization. To address the goals of PTM characterization, statistical analysis needs to summarize values in this table using appropriate statistical models, translate the goal into a model-based quantity of interest, and draw inference (i.e., characterize the uncertainty) about the quantity.

\begin{figure}[h!]
\centering
\begin{footnotesize}
\[
\begin{array}{l l c c c c c c c c c} 
\toprule
 &  & \multicolumn{4}{c}{\text{Condition }1} & \dots & \multicolumn{4}{c}{\text{Condition }I} \\ 
\cmidrule{3-6} 
\cmidrule{8-11} 
 &  &  \text{Run }1 & \text{Run }2 & \dots & \text{Run }J & \dots & \text{Run }1 & \text{Run }2 & \dots & \text{Run }J \\
\midrule
\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} & \dots & y_{1J1}^{\ast} & \dots & y_{I11}^{\ast} & y_{I21}^{\ast} & \dots & y_{IJ1}^{\ast} \\
 & \text{Feature }2 & y_{112}^{\ast} & - & \dots & y_{1J2}^{\ast} & \dots & y_{I12}^{\ast} & y_{I22}^{\ast} & \dots & y_{IJ2}^{\ast}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }K & y_{11K}^{\ast} & y_{12K}^{\ast} & \dots & y_{1JK}^{\ast} & \dots & y_{I1K}^{\ast} & y_{I2K}^{\ast} & \dots & y_{IJK}^{\ast} \\
\midrule
\multirow{4}{*}{Unmodified} & \text{Feature }1 & y_{111} & y_{121} & \dots & y_{1J1} & \dots & y_{I11} & y_{I21} & \dots & y_{IJ1} \\
 & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & - & - & \dots & -  \\
% & \text{Feature }2 & y_{112} & y_{122} & \dots & y_{1J2} & \dots & y_{I12} & y_{I22} & \dots & y_{IJ2}  \\
 & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots & \dots \\
 & \text{Feature }L & y_{11L} & y_{12L} & \dots & y_{1JL} & \dots & y_{I1L} & y_{I2L} & \dots & y_{IJL} \\
\bottomrule
\end{array}
\]
%\[
%\begin{array}{l l c c c c c} 
%\toprule
% &  & \multicolumn{2}{c}{\text{Condition }1} &  & \multicolumn{2}{c}{\text{Condition }2} \\ 
%\cmidrule{3-4} 
%\cmidrule{6-7} 
% &  &  \text{Run }1 & \text{Run }2 &  & \text{Run }1 & \text{Run }2 \\
%\midrule
%\multirow{4}{*}{\begin{turn}{90} {\text{Modified}} \end{turn}} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
%%\multirow{4}{*}{Modified} & \text{Feature }1 & y_{111} & y_{121} &  & y_{211} & y_{221} \\
% & \text{Feature }2 & y_{112} & - &  & y_{212} & y_{222} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }5 & y_{115} & y_{125} &  & y_{215} & y_{225} \\
%\midrule
%\multirow{7}{*}{\begin{turn}{90} {\text{Unmodified}} \end{turn}} & \text{Feature }1 & y_{111}^{\ast} & y_{121}^{\ast} &  & y_{211}^{\ast} & y_{221}^{\ast} \\
% & \text{Feature }2 & y_{112}^{\ast} & y_{122}^{\ast} &  & - & y_{222}^{\ast} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }7 & y_{117}^{\ast} & y_{127}^{\ast} &  & y_{217}^{\ast} & y_{227}^{\ast} \\
% & \text{Feature }8 & y_{118}^{\ast} & y_{128}^{\ast} &  & y_{218}^{\ast} & y_{228}^{\ast} \\
% & \dots & \dots & \dots &  & \dots & \dots \\
% & \text{Feature }11 & y_{1,1,11}^{\ast} & y_{1,2,11}^{\ast} &  & y_{2,1,11}^{\ast} & y_{2,2,11}^{\ast} \\
%\bottomrule
%\end{array}
%\]
\end{footnotesize}
\caption{Representation of the data of modified peptides at one site and unmodified peptides of the same protein, with $I$ conditions and $J$ replicate runs. Abundances of the PTM and protein are quantified by multiple spectral features (peptide ions, $K$ for modified peptides and $L$ for unmodified peptides). Some spectral features can be missing (shown as $-$), either randomly in individual runs or completely in certain conditions. In real practice, the number of runs can vary across conditions. \label{fig:dtable}}
\end{figure}

@@

Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i'$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. The $t$-test is typically performed based on the log of summarized value. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
\]
For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
%which is equivalent to take the difference of their log abundances. 
The quantity is denoted by $d_{ij}$ and is given by
\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
\]
Alternatively, run-level summary to be described in \secref{sec:sum} can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
\[
\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
\]
where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

@@

\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
\]

@@

\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
\]

@@

\[
\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
\]

@@

To characterize the observed feature intensities, different levels of variations are expressed using linear mixed models in consideration of the following factors in a simple experimental design: modification, condition, run, and feature. As different degrees of variability are present in the feature intensities of modified and unmodified peptides, they are expressed by separate models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Statistical modeling and inference}
%\label{sec:model}

%The observed log-intensity of a modified peptide feature is denoted by $y_{ijk}$ and represented as
%\[
%y_{ijk} = \psi + C_{i} + R_{j(i)} + F_{k} + (R \times F)_{ijk},
%\]
%where the effects of condition and feature are modeled as fixed effects: 
%\[
%\sum_{i=1}^{I} C_{i} = 0, \quad \sum_{k=1}^{K} F_{k} = 0,
%\]
%and the effects of run and its interaction with feature are considered as random effects arising from normal distribution with mean $0$: 
%\[
%R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2}), \qquad
%(R \times F)_{ijk} = \epsilon_{ijk} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon}^{2}).
%\]
%Similarly, the observed log-intensity of an unmodified peptide feature is denoted by $y_{ijl}^{\ast}$ and represented as 
%\[
%y_{ijl}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast} + F_{l}^{\ast} + (R \times F)_{ijl}^{\ast}, 
%\]
%where the effects of condition and feature are modeled as fixed effects: 
%\[
%\sum_{i=1}^{I} C_{i}^{\ast} = 0, \quad \sum_{l=1}^{L} F_{l}^{\ast} = 0,
%\]
%and 
%\[
%R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2}), \qquad
%(R \times F)_{ijl}^{\ast} = \epsilon_{ijl}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\epsilon^{\ast}}^{2}).
%\]

@@

Run-level summarization of feature intensities for each PTM site is carried out as in the sub-plot model of MSstats~\cite{choi_etal_14a}, which involves a) imputation of censored missing values, and b) summarization of feature intensities using Tukey's median polish. \cite{tukey_77a} The run-level summary for the PTM in Run $j$ of Condition $i$ is denoted by $\hat{y}_{ij}^{\ast}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Model-based inference of the underlying abundance in a simple balanced experiment}
\label{sec:infer}

In the case of a balanced design with multiple conditions and without technical replicates, the PTM abundance in each run is represented as 
\[
\hat{y}_{ij}^{\ast} = \psi^{\ast} + C_{i}^{\ast} + R_{j(i)}^{\ast},
\]
where $\sum_{i=1}^{I} C_{i}^{\ast} = 0$, $R_{j(i)}^{\ast} = \gamma_{j(i)}^{\ast} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma^{\ast}}^{2})$. 
Similarly, the protein abundance in each run is expressed as
\[
\hat{y}_{ij} = \psi + C_{i} + R_{j(i)},
\]
where $\sum_{i=1}^{I} C_{i} = 0$, $R_{j(i)} = \gamma_{j(i)} \stackrel{\text{iid}}{\sim} \mathcal{N}(0, \sigma_{\gamma}^{2})$. The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$, respectively, and the values are estimated as: 
\begin{align*}
\hat{\mu}_{i}^{\ast} &= \hat{\psi}^{\ast} + \hat{C}_{i}^{\ast} = \frac{1}{J} \hat{y}_{i+}^{\ast} \\
\hat{\mu}_{i} &= \hat{\psi} + \hat{C}_{i} = \frac{1}{J} \hat{y}_{i+},
\end{align*}
where the standard errors of the estimates are $\mathrm{SE^{\ast}}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$ and $\mathrm{SE}(\hat{\mu}_{i}) = (\hat{\sigma}_{\gamma^{\ast}}^{2} / J)^{1/2}$.
%\[
%\left( \frac{\hat{\sigma}_{\gamma}^{2}}{J} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
%\]
%\[
%\left( \frac{1}{J} \hat{\sigma}_{\gamma}^{2} \right)^{1/2}, \quad \left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{2}}{J} \right) ^ {1/2}.
%\]
Based on the estimates $\hat{\mu}_{i}^{\ast}$ and $\hat{\mu}_{i}$, the adjusted log-abundance of the PTM is given by $(\hat{\mu}_{i}^{\ast} - \hat{\mu}_{i})$ and the standard error of the estimate is 
\[
\left[ \frac{1}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
\]

@@

With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
%Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
\begin{align*}
H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
\end{align*}
In a balanced design, the log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right],
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
\]
A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.

The formulations for $\hat{\Delta}$, $SE(\hat{\Delta})$, and degrees of freedom shown above are for a simple balanced experiment. They can also be extended to more realistic experiments with complex designs. This is discussed further in Section \ref{sec:complex_methods}

@@

The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively. The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations. With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}. Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
\[
\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
\]
where 
\[
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
\]
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs. More details can be found in~\cite{oberg_vitek_09a}.

@@
